---
title: "Race and Crime Playground"
author: "Alejandro Hernandez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
subtitle: Exoneree Project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = paste(head(unlist(strsplit(getwd(), "/")), -1),
                                      collapse = "/"))
library(tidyverse)
# library(gridExtra)
library(gtsummary)
library(gt)
library(knitr)
```

```{r data-loading, echo=FALSE}
rm(list = ls()) # clear workspace
# load data
exonerees <- read.csv("data/exonerees1989-may2024.csv")


# correct Crime Category to be a factor
crime_levels <- c("Violent felony", "Misdemeanor", "Nonviolent felony", 
                  "Misconduct")
exonerees$Crime.Category <- factor(exonerees$Crime.Category, 
                                   levels = crime_levels)

# correct Age Group variables to be factors
age_levels = c("0 - 17", "18 - 29", "30 - 39", "40 - 49", "50 - 64", "65 - 100")
exonerees$Age.Group.at.Crime <- factor(exonerees$Age.Group.at.Crime, 
                                       levels = age_levels)
exonerees$Age.Group.at.Conviction <- factor(exonerees$Age.Group.at.Conviction, 
                                            levels = age_levels)
exonerees$Age.Group.at.Release <- factor(exonerees$Age.Group.at.Release, 
                                         levels = age_levels)
exonerees$Age.Group.at.Exoneration <- factor(exonerees$Age.Group.at.Exoneration,
                                             levels = age_levels)

# correct Race to be a factor
race_levels <- c("Black", "White", "Hispanic", "Asian", "Native American",
                 "Other")
exonerees$Race <- factor(exonerees$Race, levels = race_levels)

# correct Sex to be a factor
exonerees$Sex <- as.factor(exonerees$Sex)

# correct binary tags to also be factors
tag_vars <- c("F.MFE","FC","ILD","P.FA","DNA","MWID","OM","CV","IO","SA","CIU",
              "NC","P","H","JV","CDC","F","JI","M","CSH","SBS","A","FED","PH",
              "BM")
exonerees <- exonerees %>% mutate_at(tag_vars, as.factor)
```

```{r include=FALSE}
# make sure everything looks good
summary(exonerees)
dim(exonerees)
```


# Distribution Tables

Create a tables of distributions across types of offense.

```{r echo=FALSE}
# build summary tables
# 
gt_table1 <- exonerees %>% select(Crime.Category, 
                                  Sex, 
                                  Race, 
                                  Age.Group.at.Conviction) %>%
  gtsummary::tbl_summary(
    by = Crime.Category,
    statistic = list(
      all_categorical() ~ "{n} ({p}%)")) %>%   # count (percentage) for factors
  add_overall() %>%
  bold_labels() %>%
  modify_header(label ~ "**Variable**") %>%
  modify_spanning_header(c(stat_1, stat_2) ~ "**Crime Category**") %>%
  as_gt()

gt_table2 <- exonerees %>% select(Crime.Category, 
                                  Age.at.Crime, 
                                  Age.at.Conviction, 
                                  Age.at.Release, 
                                  Age.at.Exoneration) %>%
  tbl_summary(
    by = Crime.Category,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})")) %>% # mean (IQR) for continuous
  add_overall() %>%
  bold_labels() %>%     # make the labels bold
  modify_header(label ~ "**Variable**") %>%  # Customize the header
  modify_spanning_header(c(stat_1, stat_2) ~ "**Crime Category**") %>%
  as_gt()


gt_table3 <- exonerees %>% select(Crime.Category, 
                                  Months.Incarcerated, 
                                  Months.Convicted, 
                                  Years.Incarcerated, 
                                  Years.Convicted) %>%
  gtsummary::tbl_summary(
    by = Crime.Category,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})")) %>%
  add_overall() %>%
  bold_labels() %>%     # make the labels bold
  modify_header(label ~ "**Variable**") %>%  # Customize the header
  modify_spanning_header(c(stat_1, stat_2) ~ "**Crime Category**") %>%
  as_gt()

# save the gt table as a PNG file
gt::gtsave(gt_table1, filename = "images/crime-categories.png", path = getwd())
gtsave(gt_table2, filename = "images/crime-ages.png", path = getwd())
gtsave(gt_table3, filename = "images/crime-time.png", path = getwd())
```

# Distribution Graphs

First, we want to verify that, when interested in time spent wrongly imprisoned, the categories of Sex, Race, and Age are distinct groups. As we do so, we will describe how the distribution of time spent wrongly imprisoned for one group compares to its counterpart(s).

Note that time is measured here in months.

```{r distribution-of-time-spent-imprisoned, echo=FALSE}
gg_time_imprisoned <- exonerees %>%
  filter(!is.na(Months.Incarcerated)) %>%
  ggplot() +
  # jitter points (these must be overlayed by the horizontal line)
  geom_jitter(aes(y = Months.Incarcerated, x = ""), 
              width = 0.4, color = "aliceblue") +

  # marker at 1 year
  geom_hline(yintercept = 12, color = "red") +
  # time spent wrongfully imprisoned
  geom_violin(aes(y = Months.Incarcerated, x = "")) +
  geom_boxplot(aes(y = Months.Incarcerated, x = ""), alpha = 0.5) +
  
  # plot specifics
  ylab("Months spent wrongfully imprisoned") + xlab("") +
  scale_y_continuous(breaks = seq(0, 400, 120)) +
  theme_bw()

gg_time_imprisoned
```

```{r verify-covars, echo=FALSE}
gg_time_imprisoned + facet_wrap(vars(Sex))
gg_time_imprisoned + facet_wrap(vars(Race))
gg_time_imprisoned + facet_wrap(vars(Age.Group.at.Conviction))
```

# Race, adjusting for Crime

## Sample size table

```{r sample-sizes, echo=FALSE}
# race overall
exonerees %>%
  group_by(Race) %>%
  summarise(n = n())

# crime overall
exonerees %>%
  group_by(Worst.Crime.Display) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

# crime, by race
gt_table4 <- exonerees %>% 
  select(Race, Offense = Worst.Crime.Display) %>%
  tbl_summary(
    by = Race,
    statistic = list(
      all_categorical() ~ "{n} ({p}%)")) %>%   # count (percentage)
  add_overall() %>%
  bold_labels() %>%
  modify_header(label ~ "**Variable**") %>%
  modify_spanning_header(c(stat_1, stat_2) ~ "**Race**")

gt_table4_ordered <- gt_table4 %>%
  as.data.frame() %>%
  mutate(overall.count = parse_number(`**Overall**, N = 2,640`)) %>%
  arrange(desc(overall.count)) %>%
  select(-overall.count) %>% 
  head(-1) %>%    # drop the __offense__ row, which I haven't cleanly avoided
  gt()

gt_table4 <- gt_table4 %>% as_gt()

gtsave(gt_table4, filename = "images/sample-sizes.png", path = getwd())
gtsave(gt_table4_ordered, filename = "images/sample-sizes-ordered.png", path = getwd())

gt_table4_ordered

```

## Time served table

```{r years-served-table, echo=FALSE}
# overall
exonerees %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  knitr::kable(caption = "Years spent incarcerated and convicted by all exonerees")

## in (descending) order of group size:
# Black
exonerees %>% filter(Race == "Black") %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by Black exonerees")

# white
exonerees %>% 
  filter(Race == "White") %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by white exonerees")

# Hispanic
exonerees %>% filter(Race == "Hispanic") %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by Hispanic exonerees")

# Asian
exonerees %>% filter(Race == "Asian") %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by Asian exonerees")

# Native American
exonerees %>% filter(Race == "Native American") %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by Native American exonerees")

# Other
exonerees %>% filter(Race == "Other") %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by exonerees who are 
        not Black, white, Hispanic, Asian, or Native American")

# NA
exonerees %>% filter(is.na(Race)) %>%
  group_by(Offense = Worst.Crime.Display) %>%
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  arrange(desc(Incarcerated)) %>%
  kable(caption = "Years spent incarcerated and convicted by exonerees of unknown
        race")
```

## Age distributions

```{r}
plot_ages <- function (data) {
  gg <- data %>%
    filter(!is.na(Age.at.Crime)) %>%
    ggplot() +
    # jitter points (these must be overlayed by the horizontal line)
    geom_jitter(aes(y = Age.at.Crime, x = "(1) Crime"), 
                width = 0.4, color = "aliceblue", alpha = 0.4) +
    geom_jitter(aes(y = Age.at.Conviction, x = "(2) Conviction"), 
                width = 0.4, color = "aliceblue", alpha = 0.4) +
    geom_jitter(aes(y = Age.at.Release, x = "(3) Release"), 
                width = 0.4, color = "aliceblue", alpha = 0.4) +
    geom_jitter(aes(y = Age.at.Exoneration, x = "(4) Exoneration"), 
                width = 0.4, color = "aliceblue", alpha = 0.4) +
    # marker at 18 year
    geom_hline(yintercept = 18, color = "red") +
    # age at crime
    geom_violin(aes(y = Age.at.Crime, x = "(1) Crime")) +
    geom_boxplot(aes(y = Age.at.Crime, x = "(1) Crime"), alpha = 0.5) +
    # age at conviction
    geom_violin(aes(y = Age.at.Conviction, x = "(2) Conviction")) +
    geom_boxplot(aes(y = Age.at.Conviction, x = "(2) Conviction"), alpha = 0.5) +
    # age at release
    geom_violin(aes(y = Age.at.Release, x = "(3) Release")) +
    geom_boxplot(aes(y = Age.at.Release, x = "(3) Release"), alpha = 0.5) +
    # age at exoneration
    geom_violin(aes(y = Age.at.Exoneration, x = "(4) Exoneration")) +
    geom_boxplot(aes(y = Age.at.Exoneration, x = "(4) Exoneration"), alpha = 0.5)+
    # plot specifics
    ylab("Age at Event") + xlab("Event") +
    scale_y_continuous(breaks = seq(0, 100, 20)) + 
    theme_bw()
  return(gg)
}

# overall
plot_ages(exonerees)

# subset of largest race groups
exonerees %>% 
  filter(Race %in% c("Black", "White", "Hispanic")) %>%
  plot_ages() + facet_wrap(vars(Race))

## crime, by sample size
# Murder
exonerees %>%
  filter(Worst.Crime.Display == "Murder",
         Race %in% c("Black", "White", "Hispanic")) %>%
  plot_ages() + facet_wrap(vars(Race))

# Drug possession or sale
exonerees %>%
  filter(Worst.Crime.Display == "Drug Possession or Sale",
         Race %in% c("Black", "White", "Hispanic")) %>%
  plot_ages() + facet_wrap(vars(Race))

# Child sex abuse
exonerees %>%
  filter(Worst.Crime.Display == "Child Sex Abuse",
         Race %in% c("Black", "White", "Hispanic")) %>%  
  plot_ages() + facet_wrap(vars(Race))

# All but Murder
exonerees %>%
  filter(Worst.Crime.Display != "Murder",
         Race %in% c("Black", "White", "Hispanic")) %>%
  plot_ages() + facet_wrap(vars(Race))
```

## Longitudinal plots

```{r echo=FALSE}

```

## Distributions of time spent convicted and incarcerated

```{r echo=FALSE}
# overall
exonerees %>% 
  filter(!is.na(Months.Incarcerated)) %>%
  ggplot(aes(y = Months.Incarcerated, x = Race)) +
  # jitter points (these must be overlayed by the horizontal line)
  geom_jitter(width = 0.4, color = "aliceblue") +
  
  # marker at 1 year
  geom_hline(yintercept = 12, color = "red") +
  # time spent wrongfully imprisoned
  geom_violin() +
  geom_boxplot(alpha = 0.5) +
  
  # plot specifics
  ylab("Months spent wrongfully imprisoned") + xlab("") +
  scale_y_continuous(breaks = seq(0, 400, 120)) +
  theme_bw() +
  labs(
    title = "Time spent incarcerated by exonerees",
    subtitle = paste0(
    "Black: ", nrow(exonerees %>% filter(Race == "Black")), 
    " (", floor(100*nrow(exonerees %>% filter(Race == "Black")) / 
                  nrow(exonerees)), "%); ",
    "White: ", nrow(exonerees %>% filter(Race == "White")), 
    " (", floor(100*nrow(exonerees %>% filter(Race == "White")) / 
                  nrow(exonerees)), "%); ",
    "Hispanic: ", nrow(exonerees %>% filter(Race == "Hispanic")), 
    " (", floor(100*nrow(exonerees %>% filter(Race == "Hispanic")) / 
                  nrow(exonerees)), "%); ",
    "\n",
    "Asian: ", nrow(exonerees %>% filter(Race == "Asian")), 
    " (", floor(100*nrow(exonerees %>% filter(Race == "Asian")) / 
                  nrow(exonerees)), "%); ",
    "Native American: ", nrow(exonerees %>% filter(Race == "Native American")), 
    " (", floor(100*nrow(exonerees %>% filter(Race == "Native American")) / 
                  nrow(exonerees)), "%); ",
    "Other: ", nrow(exonerees %>% filter(Race == "Other")), 
    " (", floor(100*nrow(exonerees %>% filter(Race == "Other")) / 
                  nrow(exonerees)), "%); ",
    "NA: ", nrow(exonerees %>% filter(is.na(Race))), 
    " (", floor(100*nrow(exonerees %>% filter(is.na(Race))) / 
                  nrow(exonerees)), "%)"))



## crime, by total time spent incarcerated
plot_times <- function (data, crime_filter) {
  
  # filter data
  data <- data %>% 
    filter(!is.na(Months.Incarcerated),
           Worst.Crime.Display == crime_filter,
           Race %in% c("Black", "White", "Hispanic"))
  
  # plot distribution of time spent incarcerated
  gg <- data %>%
    ggplot(aes(y = Months.Incarcerated, x = Race)) +
    # jitter points (these must be overlayed by the horizontal line)
    geom_jitter(width = 0.4, color = "aliceblue") +

    # marker at 1 year
    geom_hline(yintercept = 12, color = "red") +
    # time spent wrongfully imprisoned
    geom_violin() +
    geom_boxplot(alpha = 0.5) +
    
    # plot specifics
    ylab("Months spent wrongfully imprisoned") + xlab("") +
    scale_y_continuous(breaks = seq(0, 400, 120)) +
    theme_bw() +
    labs(
    title = paste("Time spent incarcerated by exonerees of", crime_filter,
                  "convictions"),
    subtitle = paste0(
    "Black: ", nrow(data %>% filter(Race == "Black")), 
    " (", floor(100*nrow(data %>% filter(Race == "Black")) / 
                  nrow(data)), "%); ",
    "White: ", nrow(data %>% filter(Race == "White")), 
    " (", floor(100*nrow(data %>% filter(Race == "White")) / 
                  nrow(data)), "%); ",
    "Hispanic: ", nrow(data %>% filter(Race == "Hispanic")), 
    " (", floor(100*nrow(data %>% filter(Race == "Hispanic")) / 
                  nrow(data)), "%)"))
  
  return(gg)
}

# Murder
plot_times(exonerees, crime_filter = "Murder")

# Child sex abuse
plot_times(exonerees, crime_filter = "Child Sex Abuse")

# Sexual assault
plot_times(exonerees, crime_filter = "Sexual Assault")

# Robbery
plot_times(exonerees, crime_filter = "Robbery")

# Attempted murder
plot_times(exonerees, crime_filter = "Attempted Murder")

# Drug possession or sale
plot_times(exonerees, crime_filter = "Drug Possession or Sale")

# Assault
plot_times(exonerees, crime_filter = "Assault")

# Manslaughter
plot_times(exonerees, crime_filter = "Manslaughter")
```

**End of document.**
