---
title: "Univariate Exploratory Data Analysis"
subtitle: "Exoneree Project"
author: "Alejandro Hernandez"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gridExtra)
```

```{r data-loading}
rm(list = ls()) # clear workspace

# load data
exonerees <- read.csv("data/exonerees-may2024.csv")
```

# DEMOGRAPHIC

### Name

```{r name}

```

### Sex

```{r sex}

```

### Race

```{r race}

```

### Age at crime

```{r age-at-crime}

```

## Derived demographic variables

### Age at conviction

```{r age-at-conviction}

```

### Age at exoneration

```{r age-at-exoneration}

```

### Age at release

```{r age-at-release}

```

# CONVICTION

### State

```{r state}

```

### County

```{r county}

```

### Year of crime

```{r year-of-crime}

```

### (Worst) crime

```{r worst-crime}

```

### Sentence

```{r sentence}

```

### Date of 1st conviction

```{r date-conviction}

```

### Conviction tags

## Derived conviction variables

### Year of 1st conviction

```{r year-conviction}

```

# EXONERATION

### Date of exoneration

```{r date-exoneration}

```

### Date of release

```{r date-release}

```

### Date of posting

```{r date-posting}

```

### Exoneration tags

## Derived exoneration variables

### Years served

```{r years-served}

```

### Year of exoneration

```{r year-exoneration}

```

### Year of release

```{r year-release}

```

### Year of posting

```{r year-posting}

```

**End of document**
