---
title: "codebook"
author: "Alejandro Hernandez"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(gridExtra)
```

```{r}
exonerees <- read.csv("publicspreadsheet-may2024.csv",
                      na.strings = c("NA", "Don't Know", ""))

# Identify variables with missing data
colSums(is.na(exonerees))[colSums(is.na(exonerees)) != 0]
# Identify rows with missing data
exonerees[is.na(exonerees$Age),]
```

## Data Summaries

### Personal Demographics 
Name, race, age at crime, sex

```{r}
exonerees %>%
  dplyr::select(Last.Name, First.Name, Race, Age, Sex) %>%
  dplyr::mutate_if(is.character, as.factor) %>%
  summary
```

### Details of the Crime 
State, county, year, criminal charge, and tags indicating factors relating to charges (e.g., co-defendant confession, federal prosecution, child abuse, plea deal, etc.). 

```{r}
exonerees %>%
  dplyr::select(State, County, Date.of.Crime.Year, Worst.Crime.Display) %>%
  dplyr::mutate_all(as.factor) %>%
  summary

# count number of federal cases with only one federal district in that state,
# such that County == ""
# exonerees %>%
  # select(County) %>% mutate_at(c("County"), as.factor) %>%
  # filter(County == "") %>%
  # nrow
```

### Details of the Exoneration 
Sentence, date convicted, date released, and tags indicating factors relating to evidence (e.g., false confession, mistaken witness ID, false or misleading forensic evidence, inadequate legal defense, etc.).

```{r}
exonerees %>%
  dplyr::select(Sentence, Date.of.1st.Conviction, Date.of.Release, 
                Date.of.Exoneration) %>%
  dplyr::mutate_at(c("Sentence"), as.factor) %>%
  summary
```
### Tags

```{r}
tags_list <- exonerees %>%
  dplyr::select(Tags) %>%
  dplyr::mutate(Tags = strsplit(Tags, ";#"))

max_tags <- max(sapply(tags_list$Tags, length))

df_list_padded <- lapply(tags_list$Tags, function(x) c(x, rep(NA, max_tags - length(x))))

df_padded <- as.data.frame(do.call(rbind, df_list_padded))

head(df_padded)
colnames(df_padded) <- paste0("tag_", 1:max_tags)

```
```{r}
# plot of number of exonerees over year by race
exonerees <- exonerees %>%
  mutate(Date.of.Crime.Year = gsub(",", "", Date.of.Crime.Year),
         Date.of.Crime.Year = as.numeric(Date.of.Crime.Year))

# plot of number of exonerees over year by race
exonerees %>%
  filter(Date.of.Crime.Year >= 1989) %>%
  group_by(Race, Date.of.Crime.Year) %>%
  summarize(Count = n()) %>%
  ggplot(aes(x = Date.of.Crime.Year, y = Count, color = Race)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    scale_x_continuous(
      breaks = seq(min(exonerees$Date.of.Crime.Year, na.rm = TRUE),
                   max(exonerees$Date.of.Crime.Year, na.rm = TRUE),
                                    by = 5)) +
    theme_bw()


# plot of number of exonerees over age by race
exonerees %>%
  filter(Date.of.Crime.Year >= 1989) %>%
  group_by(Race, Age) %>%
  summarize(Count = n()) %>%
  ggplot(aes(x = Age, y = Count, color = Race)) +
    geom_point(alpha = .4) +
    geom_smooth(se = FALSE) +
    scale_x_continuous(
      breaks = seq(min(exonerees$Age, na.rm = TRUE),
                   max(exonerees$Age, na.rm = TRUE),
                                    by = 5)) +
    theme_bw()

# boxplot of age by race, faceted by sex 
exonerees %>%
  filter(Date.of.Crime.Year >= 1989) %>%
  group_by(Race, Age, Sex) %>%
  summarize(Count = n()) %>%
  ggplot(aes(y = Age, x = Race, color = Race)) +
    geom_boxplot() +
    geom_jitter(width = 0.1) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(vars(Sex))


worst.crime.by.race <- exonerees %>%
  filter(Date.of.Crime.Year >= 1989) %>%
  mutate(Worst.Crime.Display = as.factor(Worst.Crime.Display)) %>%
  group_by(Race, Worst.Crime.Display) %>%
  summarize(Count = n())

# Ordered by count
worst.crime.by.race %>%
  arrange(desc(Count)) %>%
  filter(Count >= 50)

# Ordered by race
worst.crime.by.race %>%
  arrange(Race, desc(Count)) %>%
  filter(Count >= 30)

# Ordered by crime
worst.crime.by.race %>%
  arrange(Worst.Crime.Display, desc(Count)) %>%
  select(Worst.Crime.Display, everything()) %>%
  filter(Count >= 30)

# table of sentence by race
exonerees %>%
  dplyr::mutate_at(c("Sentence"), as.factor) %>%
  group_by(Race, Sentence) %>%
  summarize(Count = n()) %>%
  arrange(Sentence, desc(Count)) %>%
  filter(Count > 50)
  

exonerees %>%
  group_by(State) %>%
  summarize(Count = n()) %>%
  filter(Count >= 50) %>%
  arrange(desc(Count))

exonerees %>%
  group_by(State, Race) %>%
  summarize(Count = n()) %>%
  ungroup(Race) %>%
  mutate(Total = sum(Count),
         Prop = round(Count/sum(Count), 2)) %>%
  filter(Count >= 50) %>%
  arrange(desc(Total), State, desc(Prop)) %>%
  select(Total, everything())
```



