---
title: Race and Crime
author: Alejandro D. Hernandez
date: August 2024
output:
  pdf_document: default
subtitle: Analysis of the U.S. National Registry of Exonerations
---

```{r setup, include=FALSE}
### -----------------------------------------------------------
### Setup
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
knitr::opts_knit$set(root.dir = paste(head(unlist(strsplit(getwd(), "/")), -1),
                                      collapse = "/"))
# options(knitr.kable.NA = '-')
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "allcode")]

# import required libraries
library(tidyverse)
library(gridExtra)
library(gtsummary)
library(gt)
library(knitr)
library(kableExtra)
```

# Background

The pursuit of justice is a fundamental pillar of the United States legal system, aiming to hold accountability for the guilty and protection of the innocent. However, the justice system design induces wrongful convictions through eyewitness misidentification, false or misleading forensic evidence, perjury, official misconduct, and inadequate legal defense. The consequences of wrongful conviction are serious and often irreversible, resulting in loss of liberty and damage to the reputation of individuals and their families. Increased scrutiny of criminal cases and efforts of dedicated legal advocates have facilitated the *exoneration* (reversal of conviction and clearing of charges) of numerous individuals who were unjustly convicted. Exoneration brings attention to shortcomings of the criminal justice system and consideration of innocents who are currently imprisoned.

The National Registry of Exonerations is a project from the University of California, Irvine; University of Michigan; and Michigan State University which gathers detailed information on every exoneration in the United States since 1989. Their extensive registry includes personal demographics and describes criminal sentences, time served, and pathway to exoneration. This wealth of information allows researchers, policymakers, legal practitioners, and the public to analyze trends, identify systemic issues within the criminal justice system, and advocate for reforms aimed at preventing future wrongful convictions.

# Objective

In this study, we examine the National Registry of Exonerations from the University of Michigan to analyze individuals exonerated in the United States between 1989 and 2024 through descriptive statistics and statistical modeling. This document and our work are in progress.

```{r data-loading}
### -----------------------------------------------------------
### Data Loading

# clear workspace
rm(list = ls())

# load data
exonerees <- read.csv("data/exonerees1989-may2024.csv")

# correct Crime Category to be a factor
crime_levels <- c("Violent felony", "Misdemeanor", "Nonviolent felony", 
                  "Misconduct")
exonerees$Crime.Category <- factor(exonerees$Crime.Category, 
                                   levels = crime_levels)

# correct Age Group variables to be factors
age_levels = c("0 - 17", "18 - 29", "30 - 39", "40 - 49", "50 - 64", "65 - 100")
exonerees$Age.Group.at.Crime <- factor(exonerees$Age.Group.at.Crime, 
                                       levels = age_levels)
exonerees$Age.Group.at.Conviction <- factor(exonerees$Age.Group.at.Conviction, 
                                       levels = age_levels)
exonerees$Age.Group.at.Release <- factor(exonerees$Age.Group.at.Release, 
                                       levels = age_levels)
exonerees$Age.Group.at.Exoneration <- factor(exonerees$Age.Group.at.Exoneration,
                                       levels = age_levels)

# correct Race to be a factor
race_levels <- c("Black", "White", "Hispanic", "Asian", "Native American",
                 "Other")
exonerees$Race <- factor(exonerees$Race, levels = race_levels)

# correct Sex to be a factor
exonerees$Sex <- as.factor(exonerees$Sex)

# correct binary tags to also be factors
tag_vars <- c("F.MFE","FC","ILD","P.FA","DNA","MWID","OM","CV","IO","SA","CIU",
              "NC","P","H","JV","CDC","F","JI","M","CSH","SBS","A","FED","PH",
              "BM")
exonerees <- exonerees %>% mutate_at(tag_vars, as.factor)
```

```{r, include=FALSE}
# make sure everything looks good
summary(exonerees)
dim(exonerees)
```

### Which offenses do we examine?

```{r ordered-sample-sizes-table}
### -----------------------------------------------------------
### Ordered Sample Size Tables
exonerees %>%
  select(Race, Worst.Crime.Display) %>%
  filter(!is.na(Race)) %>%
  # ordered frequency of offenses, overall and by race
  tbl_summary(
    by = Race,
    statistic = list(
      all_categorical() ~ "{n} ({p}%)")) %>%   # count (percent)
  add_overall() %>%
  bold_labels() %>%
  modify_header(label ~ "Offense") %>%
  modify_spanning_header(c(stat_1, stat_2) ~ "Race") %>%
  # convert to a data frame and arrange by a new overall count column
  as.data.frame() %>%
  mutate(overall.count = parse_number(`**Overall**, N = 2,640`)) %>%
  arrange(desc(overall.count)) %>%
  select(-overall.count) %>%  # remove the overall count column
  # only present the 6 most frequent offenses
  head() %>% 
  # table styling 
  knitr::kable(
    format = "simple", 
    caption = "List of high-frequency offenses, by race")
```

Table 1 shows the highest-frequency criminal offenses across groups of race. For example, we read the top-left-most cell as reporting that, among the 2,640 individuals in the National Registry of Exonerations whose alleged crime occurred during or after 1989, 875 or 33% were exonerated from a Murder conviction. Percents are column-wise.

```{r time-served-table}
### -----------------------------------------------------------
### Time Served Table

# write column headers now, for later use
header_Incarcerated <- paste0(
  "Incarcerated (", 
  formatC(sum(exonerees$Years.Incarcerated), big.mark = ","),
  " total)")
header_Convicted <- paste0(
  "Convicted (", 
  formatC(sum(exonerees$Years.Convicted), big.mark = ","),
  " total)")

# time served incarcerated and convicted (ordered), by offense
exonerees %>%
  group_by(Offense = Worst.Crime.Display) %>%
  # summarize total years
  summarise(Incarcerated = sum(Years.Incarcerated),
            Convicted = sum(Years.Convicted)) %>%
  # calculate percents
  mutate(Incarcerated.perc = round(100*Incarcerated / sum(Incarcerated)),
         Convicted.perc = round(100*Convicted / sum(Convicted))) %>%
  # arrange and filter out offenses that collectively have not served 90+ years
  arrange(desc(Incarcerated)) %>%
  filter(Incarcerated > 90) %>%
  # # reformat count to character
  mutate(Incarcerated = formatC(Incarcerated, big.mark = ","),
         Convicted = formatC(Convicted, big.mark = ",")) %>%
  # combine count and percent
  mutate(Incarcerated = paste0(Incarcerated, " (", Incarcerated.perc, "%)"),
         Convicted = paste0(Convicted, " (", Convicted.perc, "%)")) %>%
  # remove unused columns
  select(-contains(".perc")) %>%
  # table styling 
  knitr::kable(
    format = "simple",
    # rename header to include total time across all offenses
    col.names = c("Offense", header_Incarcerated, header_Convicted),
    caption = "Total years spent incarcerated and convicted, by offense")
```

Table 2 shows the criminal offenses that account for the most time spent wrongfully incarcerated/convicted. For example, we read the top-left-most cell as reporting that, among the 17,086 years spent wrongfully incarcerated across all individuals in the National Registry of Exonerations whose alleged crime occurred during or after 1989, 10,979 or 64% were due to a Murder conviction. Again, percents are column-wise.

Table 2 was reproduced for each group of race (see Supplementary), which found that this order of offenses accounting for the most years lost to incarceration or conviction are similar across all subsets of races, especially the largest.

**These tables motivate us to compare the experience of people convicted of these popular and severe offenses between groups of race, so as to measure the extent to which sentence/incarceration is associated with race, within the National Registry of Exonerations.**

\pagebreak

# I. Distribution of time spent incarcerated and convicted, by offense and race

```{r function-plot-times}
### -----------------------------------------------------------
### Distribution of time spent incarcerated and convicted, by offense and race
## crime, by total time spent incarcerated
plot_times <- function (data, crime_filter = NA) {
  
  ## filter the data by crime (optional)
  if (is.na(crime_filter)) {
    crime_filter <- "all" # this is used later for labeling
  } else {
    data <- data %>% 
      filter(Worst.Crime.Display == crime_filter) %>%
      filter(!is.na(Months.Incarcerated) & !is.na(Months.Convicted))
  }
  
  ## only plot groups of race with 10 or more members present
  # get counts
  counts <- data %>%
    group_by(Race) %>%
    summarise(n = n()) %>%
    mutate(prop = n / sum(n)) %>%
    filter(n >= 10)
  # filter race groups with too few members
  plot_data <- data %>% filter(Race %in% counts$Race)
  
  ## write the subtitle of the plot, which shows counts of race groups
  gg_subtitle <- ""
  included_races <- unique(plot_data$Race)
  for (i in 1:length(included_races)) {
    race <- race_levels[i]
    count <- nrow(plot_data %>% filter(Race == race))
    gg_subtitle <- paste0(
      gg_subtitle, # previous groups
      # title, count, and percent (from data with all race groups)
      # ensure percent is over 'data' not the race filtered 'plot_data'
      race, ": ", count, " (", floor(100*count/nrow(data)), "%)")
    # add a semi-colon if the list isn't done
    if(i < length(included_races))
      gg_subtitle <- paste0(gg_subtitle, "; ")
    # add a new line after the fourth race group
    if (i == 4)
      gg_subtitle <- paste0(gg_subtitle, "\n")
    }
  
  ## plot distribution of time spent incarcerated
  gg_incarated <- plot_data %>%
    ggplot(aes(y = Months.Incarcerated, x = Race)) +
    # jitter points (these must be overlayed by the horizontal line)
    geom_jitter(width = 0.4, color = "aliceblue") +
    # marker at 1 year
    geom_hline(yintercept = 12, color = "red") +
    # time spent wrongfully imprisoned
    geom_violin() +
    geom_boxplot(alpha = 0.5) +
    # plot specifics
    scale_y_continuous(breaks = seq(0, 400, 120)) +
    theme_bw() +
    labs(
      x = "",
      y = "Months imprisoned",
      title = paste("Time spent incarcerated by exonerees of", 
                    crime_filter, "convictions"),
      subtitle = gg_subtitle
      )
  
  ## plot distribution of time spent convicted
  gg_convicted <- plot_data %>%
    ggplot(aes(y = Months.Convicted, x = Race)) +
    # jitter points (these must be overlayed by the horizontal line)
    geom_jitter(width = 0.4, color = "aliceblue") +
    # marker at 1 year
    geom_hline(yintercept = 12, color = "red") +
    # time spent wrongfully convicted
    geom_violin() +
    geom_boxplot(alpha = 0.5) +
    # plot specifics
    scale_y_continuous(breaks = seq(0, 400, 120)) +
    theme_bw() +
    labs(
      x = "",
      y = "Months convicted",
      title = paste("Time spent convicted by exonerees of", 
                    crime_filter, "convictions"),
      subtitle = gg_subtitle
      )
  
  ## tabular distribution of time spent convicted
  tbl_convicted <- data %>%
    # calculate summary statistics by race
    group_by(across(Race)) %>%
    summarize(
      N = n(),
      Sum = sum(Months.Convicted),
      Quartile25 = quantile(Months.Convicted, 0.25, na.rm = TRUE),
      Quartile50 = median(Months.Convicted, na.rm = TRUE),
      Quartile75 = quantile(Months.Convicted, 0.75, na.rm = TRUE),
      Mean = mean(Months.Convicted, na.rm = TRUE),
      SD = sd(Months.Convicted, na.rm = TRUE)
    ) %>%
    # append a last row for summary statistics, regardless of race 
    add_row(
      data %>%
        summarize(
          Race = "All",
          N = n(),
          Sum = sum(Months.Convicted, na.rm = TRUE),
          Quartile25 = quantile(Months.Convicted, 0.25, na.rm = TRUE),
          Quartile50 = median(Months.Convicted, na.rm = TRUE),
          Quartile75 = quantile(Months.Convicted, 0.75, na.rm = TRUE),
          Mean = mean(Months.Convicted, na.rm = TRUE),
          SD = sd(Months.Convicted, na.rm = TRUE)
          )
      ) %>%
    # arrange by sample size (beginning with overall)
    arrange(desc(N)) %>%
    # table styling
    knitr::kable(
      format = "latex", 
      booktabs = TRUE,
      format.args = list(big.mark = ","),
      digits = 0,
      caption = paste("Months spent convicted by exonerees of", crime_filter, 
                      "convictions")
      ) %>%
    # bold overall row
    kableExtra::row_spec(1, background = "#D3D3D3") %>%
    kableExtra::kable_styling(latex_options = "hold_position")
  
  
  ## tabular distribution of time spent incarcerated
  tbl_incarcerated <- data %>%
    # calculate summary statistics by race
    group_by(across(Race)) %>%
    summarize(
      N = n(),
      Sum = sum(Months.Incarcerated),
      Quartile25 = quantile(Months.Incarcerated, 0.25, na.rm = TRUE),
      Quartile50 = median(Months.Incarcerated, na.rm = TRUE),
      Quartile75 = quantile(Months.Incarcerated, 0.75, na.rm = TRUE),
      Mean = mean(Months.Incarcerated, na.rm = TRUE),
      SD = sd(Months.Incarcerated, na.rm = TRUE)
    ) %>%
    # append a last row for summary statistics, regardless of race 
    add_row(
      data %>%
        summarize(
          Race = "All",
          N = n(),
          Sum = sum(Months.Incarcerated, na.rm = TRUE),
          Quartile25 = quantile(Months.Incarcerated, 0.25, na.rm = TRUE),
          Quartile50 = median(Months.Incarcerated, na.rm = TRUE),
          Quartile75 = quantile(Months.Incarcerated, 0.75, na.rm = TRUE),
          Mean = mean(Months.Incarcerated, na.rm = TRUE),
          SD = sd(Months.Incarcerated, na.rm = TRUE)
          )
      ) %>%
    # arrange by sample size (beginning with overall)
    arrange(desc(N)) %>%
    # table styling
    knitr::kable(
      format = "latex", 
      booktabs = TRUE,
      format.args = list(big.mark = ","),
      digits = 0,
      caption = paste("Months spent incarcerated by exonerees of", crime_filter, 
                      "convictions")
      ) %>%
    # bold overall row
    kableExtra::row_spec(1, background = "#D3D3D3") %>%
    kableExtra::kable_styling(latex_options = "hold_position")

  
  ## return plots and tables
  return(list(
    plot_conviction = gg_convicted,
    plot_incarceration = gg_incarated,
    table_convicted = tbl_convicted,
    table_incarcerated = tbl_incarcerated)
    )
  }
```

## Overall

```{r overall-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
## Overall stats
overall_stats <- plot_times(exonerees)
# plots
overall_stats <- plot_times(exonerees)
gridExtra::grid.arrange(overall_stats$plot_incarceration,
                        overall_stats$plot_conviction + labs(subtitle = NULL))
# tables
overall_stats$table_convicted
overall_stats$table_incarcerated
```

\pagebreak

## Murder

```{r murder-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
## Murder stats
mdr_stats <- plot_times(exonerees, crime_filter = "Murder")
# plots
gridExtra::grid.arrange(mdr_stats$plot_incarceration,
                        mdr_stats$plot_conviction + labs(subtitle = NULL))
# tables
mdr_stats$table_convicted
mdr_stats$table_incarcerated
```

\pagebreak

## Child sex abuse

```{r child-sex-abuse-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Child sex abuse
csa_stats <- plot_times(exonerees, crime_filter = "Child Sex Abuse")
# plots
gridExtra::grid.arrange(csa_stats$plot_incarceration,
                        csa_stats$plot_conviction + labs(subtitle = NULL))
# tables
csa_stats$table_convicted
csa_stats$table_incarcerated
```

\pagebreak

## Sexual assault

```{r sexual-assault-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Sexual assault
sxa_stats <- plot_times(exonerees, crime_filter = "Sexual Assault")
# plots
gridExtra::grid.arrange(sxa_stats$plot_incarceration,
                        sxa_stats$plot_conviction + labs(subtitle = NULL))
# tables
sxa_stats$table_convicted
sxa_stats$table_incarcerated
```

\pagebreak

## Robbery

```{r robbery-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Robbery
rby_stats <- plot_times(exonerees, crime_filter = "Robbery")
# plots
gridExtra::grid.arrange(rby_stats$plot_incarceration,
                        rby_stats$plot_conviction + labs(subtitle = NULL))
# tables
rby_stats$table_convicted
rby_stats$table_incarcerated
```

\pagebreak

## Attempted murder

```{r attempted-murder-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Attempted murder
atm_stats <- plot_times(exonerees, crime_filter = "Attempted Murder")
# plots
gridExtra::grid.arrange(atm_stats$plot_incarceration,
                        atm_stats$plot_conviction + labs(subtitle = NULL))
# tables
atm_stats$table_convicted
atm_stats$table_incarcerated
```

\pagebreak

## Drug possession or sale

```{r drug-possession-or-sale-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Drug possession or sale
dps_stats <- plot_times(exonerees, crime_filter = "Drug Possession or Sale")
# plots
gridExtra::grid.arrange(dps_stats$plot_incarceration,
                        dps_stats$plot_conviction + labs(subtitle = NULL))
# tables
dps_stats$table_convicted
dps_stats$table_incarcerated
```

\pagebreak

## Assault

```{r assault-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Assault
asa_stats <- plot_times(exonerees, crime_filter = "Assault")
# plots
gridExtra::grid.arrange(asa_stats$plot_incarceration,
                        asa_stats$plot_conviction + labs(subtitle = NULL))
# tables
asa_stats$table_convicted
asa_stats$table_incarcerated
```

\pagebreak

## Manslaughter

```{r manslaughter-stats, fig.align='center', fig.height=6}
### -----------------------------------------------------------
# Manslaughter
msl_stats <- plot_times(exonerees, crime_filter = "Manslaughter")
# plots
gridExtra::grid.arrange(msl_stats$plot_incarceration,
                        msl_stats$plot_conviction + labs(subtitle = NULL))
# tables
msl_stats$table_convicted
msl_stats$table_incarcerated
```

\pagebreak

# II. Distribution of age at events, by offense and race

\pagebreak

# Supplementary

#### i. Which offenses account for the most time spent wrongfully incarcerated and convicted for each race? \newline

Table 2 has been recreated to show how the criminal offenses that account for the most time spent wrongfully incarcerated/convicted compare between groups of race.

```{r supp-time-served-table}
### -----------------------------------------------------------
### Supplementary Time Served Table

time_served_table <- function (race_filter) {
  # filter exonerees by race
  data <- exonerees %>% filter(Race == race_filter)
  
  # write column headers now, for later use
  header_Incarcerated <- paste0("Incarcerated (",
                         formatC(sum(data$Years.Incarcerated), big.mark = ","),
                         " total)")
  header_Convicted <- paste0("Convicted (",
                         formatC(sum(data$Years.Convicted), big.mark = ","),
                        " total)")
  
  # time served incarcerated and convicted (ordered), by offense
  table <- data %>%
    group_by(Offense = Worst.Crime.Display) %>%
    # summarize total years
    summarise(Incarcerated = sum(Years.Incarcerated),
              Convicted = sum(Years.Convicted)) %>%
    # calculate percents
    mutate(Incarcerated.perc = round(100*Incarcerated / sum(Incarcerated)),
           Convicted.perc = round(100*Convicted / sum(Convicted))) %>%
    # arrange by total years spent incarcerated 
    arrange(desc(Incarcerated)) %>%
    # reformat count to character
    mutate(Incarcerated = formatC(Incarcerated, big.mark = ","),
           Convicted = formatC(Convicted, big.mark = ",")) %>%
    # combine count and percent
    mutate(Incarcerated = paste0(Incarcerated, " (", Incarcerated.perc, "%)"),
           Convicted = paste0(Convicted, " (", Convicted.perc, "%)")) %>%
    # remove unused columns
    select(-contains(".perc")) %>%
    # select top 5
    head(5) %>%
    # table styling 
    knitr::kable(
      format = "simple",
      # rename header to include total time across all offenses
      col.names = c("Offense", header_Incarcerated, header_Convicted),
      caption = paste("Total years spent incarcerated and convicted by",
                      race_filter, "exonerees"))
  return(table)
}

# iterate over all groups of race
time_served_table("Black")
time_served_table("White")
time_served_table("Hispanic")
time_served_table("Asian")
time_served_table("Native American")
time_served_table("Other")
```

\pagebreak

#### ii. How does the distribution of time spent incarcerated/convicted for a particular class of offense compare between races?

```{r supp-plot-times, fig.align='center', fig.height=6}
### -----------------------------------------------------------
### Distribution of time spent incarcerated and convicted, by class of offense 
### and race, by total time spent incarcerated
plot_times2 <- function (data, crime_filter = NA) {
  
  ## filter the data by crime (optional)
  if (is.na(crime_filter)) {
    crime_filter <- "all" # this is used later for labeling
  } else {
    data <- data %>% 
      filter(Crime.Category == crime_filter) %>%
      filter(!is.na(Months.Incarcerated) & !is.na(Months.Convicted))
  }
  
  ## only plot groups of race with 10 or more members present
  # get counts
  counts <- data %>%
    group_by(Race) %>%
    summarise(n = n()) %>%
    mutate(prop = n / sum(n)) %>%
    filter(n >= 10)
  # filter race groups with too few members
  plot_data <- data %>% filter(Race %in% counts$Race)
  
  ## write the subtitle of the plot, which shows counts of race groups
  gg_subtitle <- ""
  included_races <- unique(plot_data$Race)
  for (i in 1:length(included_races)) {
    race <- race_levels[i]
    count <- nrow(plot_data %>% filter(Race == race))
    gg_subtitle <- paste0(
      gg_subtitle, # previous groups
      # title, count, and percent (from data with all race groups)
      # ensure percent is over 'data' not the race filtered 'plot_data'
      race, ": ", count, " (", floor(100*count/nrow(data)), "%)")
    # add a semi-colon if the list isn't done
    if(i < length(included_races))
      gg_subtitle <- paste0(gg_subtitle, "; ")
    # add a new line after the fourth race group
    if (i == 4)
      gg_subtitle <- paste0(gg_subtitle, "\n")
    }
  
  ## plot distribution of time spent incarcerated
  gg_incarated <- plot_data %>%
    ggplot(aes(y = Months.Incarcerated, x = Race)) +
    # jitter points (these must be overlayed by the horizontal line)
    geom_jitter(width = 0.4, color = "aliceblue") +
    # marker at 1 year
    geom_hline(yintercept = 12, color = "red") +
    # time spent wrongfully imprisoned
    geom_violin() +
    geom_boxplot(alpha = 0.5) +
    # plot specifics
    scale_y_continuous(breaks = seq(0, 400, 120)) +
    theme_bw() +
    labs(
      x = "",
      y = "Months imprisoned",
      title = paste("Time spent incarcerated by exonerees of", 
                    crime_filter, "convictions"),
      subtitle = gg_subtitle
      )
  
  ## plot distribution of time spent convicted
  gg_convicted <- plot_data %>%
    ggplot(aes(y = Months.Convicted, x = Race)) +
    # jitter points (these must be overlayed by the horizontal line)
    geom_jitter(width = 0.4, color = "aliceblue") +
    # marker at 1 year
    geom_hline(yintercept = 12, color = "red") +
    # time spent wrongfully convicted
    geom_violin() +
    geom_boxplot(alpha = 0.5) +
    # plot specifics
    scale_y_continuous(breaks = seq(0, 400, 120)) +
    theme_bw() +
    labs(
      x = "",
      y = "Months convicted",
      title = paste("Time spent convicted by exonerees of", 
                    crime_filter, "convictions"),
      subtitle = gg_subtitle
      )
  
  ## tabular distribution of time spent convicted
  tbl_convicted <- data %>%
    # calculate summary statistics by race
    group_by(across(Race)) %>%
    summarize(
      N = n(),
      Sum = sum(Months.Convicted),
      Quartile25 = quantile(Months.Convicted, 0.25, na.rm = TRUE),
      Quartile50 = median(Months.Convicted, na.rm = TRUE),
      Quartile75 = quantile(Months.Convicted, 0.75, na.rm = TRUE),
      Mean = mean(Months.Convicted, na.rm = TRUE),
      SD = sd(Months.Convicted, na.rm = TRUE)
      ) %>%
    # append a last row for summary statistics, regardless of race 
    add_row(
      data %>%
        summarize(
          Race = "All",
          N = n(),
          Sum = sum(Months.Convicted, na.rm = TRUE),
          Quartile25 = quantile(Months.Convicted, 0.25, na.rm = TRUE),
          Quartile50 = median(Months.Convicted, na.rm = TRUE),
          Quartile75 = quantile(Months.Convicted, 0.75, na.rm = TRUE),
          Mean = mean(Months.Convicted, na.rm = TRUE),
          SD = sd(Months.Convicted, na.rm = TRUE)
          )
      ) %>%
    # arrange by sample size (beginning with overall)
    arrange(desc(N)) %>%
    # table styling
    knitr::kable(
      format = "latex", 
      booktabs = TRUE,
      format.args = list(big.mark = ","),
      digits = 0,
      caption = paste("Months spent convicted by exonerees of", crime_filter,
                      "convictions")
      ) %>%
    # bold overall row
    kableExtra::row_spec(1, background = "#D3D3D3") %>%
    kableExtra::kable_styling(latex_options = "hold_position")
  
  
  
  ## tabular distribution of time spent incarcerated
  tbl_incarcerated <- data %>%
    # calculate summary statistics by race
    group_by(across(Race)) %>%
    summarize(
      N = n(),
      Sum = sum(Months.Incarcerated),
      Quartile25 = quantile(Months.Incarcerated, 0.25, na.rm = TRUE),
      Quartile50 = median(Months.Incarcerated, na.rm = TRUE),
      Quartile75 = quantile(Months.Incarcerated, 0.75, na.rm = TRUE),
      Mean = mean(Months.Incarcerated, na.rm = TRUE),
      SD = sd(Months.Incarcerated, na.rm = TRUE)
    ) %>%
    # append a last row for summary statistics, regardless of race 
    add_row(
      data %>%
        summarize(
          Race = "All",
          N = n(),
          Sum = sum(Months.Incarcerated, na.rm = TRUE),
          Quartile25 = quantile(Months.Incarcerated, 0.25, na.rm = TRUE),
          Quartile50 = median(Months.Incarcerated, na.rm = TRUE),
          Quartile75 = quantile(Months.Incarcerated, 0.75, na.rm = TRUE),
          Mean = mean(Months.Incarcerated, na.rm = TRUE),
          SD = sd(Months.Incarcerated, na.rm = TRUE)
          )
      ) %>%
    # arrange by sample size (beginning with overall)
    arrange(desc(N)) %>%
    # table styling
    knitr::kable(
      format = "latex", 
      booktabs = TRUE,
      format.args = list(big.mark = ","),
      digits = 0,
      caption = paste("Months spent incarcerated by exonerees of", crime_filter, 
                      "convictions")
      ) %>%
    # bold overall row
    kableExtra::row_spec(1, background = "#D3D3D3") %>%
    kableExtra::kable_styling(latex_options = "hold_position")

  
  ## return plots and tables
  return(list(
    plot_conviction = gg_convicted,
    plot_incarceration = gg_incarated,
    table_convicted = tbl_convicted,
    table_incarcerated = tbl_incarcerated)
    )
  }
```

### Violent felony

```{r, fig.align='center', fig.height=6}
## Violent felony
vlf_stats <- plot_times2(exonerees, crime_filter = "Violent felony")
# plots
gridExtra::grid.arrange(vlf_stats$plot_incarceration,
                        vlf_stats$plot_conviction + labs(subtitle = NULL))
# tables
vlf_stats$table_convicted
vlf_stats$table_incarcerated
```

\newpage

### Misdemeanor

```{r, fig.align='center', fig.height=6}
## Misdemeanor
msd_stats <- plot_times2(exonerees, crime_filter = "Misdemeanor")
# plots
gridExtra::grid.arrange(msd_stats$plot_incarceration,
                        msd_stats$plot_conviction + labs(subtitle = NULL))
# tables
msd_stats$table_convicted
msd_stats$table_incarcerated
```

\newpage

### Nonviolent felony

```{r, fig.align='center', fig.height=6}
## Nonviolent felony
nvf_stats <- plot_times2(exonerees, crime_filter = "Nonviolent felony")
# plots
gridExtra::grid.arrange(nvf_stats$plot_incarceration,
                        nvf_stats$plot_conviction + labs(subtitle = NULL))
# tables
nvf_stats$table_convicted
nvf_stats$table_incarcerated
```

### Misconduct

This class of offense had zero months of times spent incarcerated or convicted. This doesn't necessarily mean zero time, only zero complete months, because our current variable for time is measured in months, as an integer.

**End of report.**

\pagebreak

## Code Appendix

```{r allcode, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```

**End of document**
