---
title: "Univariate Exploratory Data Analysis"
author: "Alejandro Hernandez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
subtitle: Exoneree Project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = paste(head(unlist(strsplit(getwd(), "/")), -1), 
                                       collapse = "/"))
library(tidyverse)
# library(gridExtra)
# library(gtsummary)
# library(gt)
```

```{r data-loading}
rm(list = ls()) # clear workspace
# load data
exonerees <- read.csv("data/exonerees1989-may2024.csv")

# correct Crime Category to be a factor
exonerees$Crime.Category <- as.factor(exonerees$Crime.Category)
levels(exonerees$Crime.Category) <- c("Violent felony", "Misdemeanor",
                                      "Nonviolent felony", "Misconduct")

# correct Age Group variables to be factors
exonerees <- exonerees %>%
  mutate_at(vars(Age.Group.at.Crime, Age.Group.at.Conviction, 
                 Age.Group.at.Release, Age.Group.at.Exoneration), 
            as.factor)
age_levels <- c("Adult", "Middle age", "Minor", "Senior", "Young adult")
levels(exonerees$Age.Group.at.Crime) <- age_levels
levels(exonerees$Age.Group.at.Conviction) <- age_levels
levels(exonerees$Age.Group.at.Release) <- age_levels
levels(exonerees$Age.Group.at.Exoneration) <- age_levels

# correct Race to be a factor
exonerees$Race <- as.factor(exonerees$Race)
levels(exonerees$Race) <- c("Black", "White", "Hispanic", "Native American", 
                            "Asian", "Other")

# correct binary tags to also be factors
tag_vars <- c("F.MFE","FC","ILD","P.FA","DNA","MWID","OM","CV","IO","SA","CIU",
              "NC","P","H","JV","CDC","F","JI","M","CSH","SBS","A","FED","PH",
              "BM")
exonerees <- exonerees %>% mutate_at(tag_vars, as.factor)

# find negative values of Months Spent Incarcerated
exonerees %>% 
  select(ID, First.Name, Last.Name, Year.of.1st.Conviction, Year.of.Release,
         Months.Incarcerated) %>%
  filter(Months.Incarcerated < 0)

# Anthony Gayles (ID 1108) was marked as being convicted at age 49 and released 
# at 48. We elect to ignore this individual from our analysis.
exonerees <- exonerees %>% filter(ID != 1108)
```


```{r table-1}
# get columns for Table 1
exonerees_tab1 <- exonerees %>% select(
  # demographics
  Race, Sex, 
  # crime details
  Crime.Category,
  # time served
  Time.Incarcerated, Years.Incarcerated, Years.Convicted,
  # age at events
  Age.at.Exoneration, Age.Group.at.Exoneration)


# build classic Table 1
gt_table1 <- exonerees_tab1 %>%
  gtsummary::tbl_summary(
    by = Sex,
    statistic = list(
      all_continuous() ~ "{mean} ({p25}, {p75})", # mean (IQR) for continuous
      all_categorical() ~ "{n} ({p}%)")) %>%   # count (percentage) for factors
  add_overall() %>%
  bold_labels() %>%     # makes the labels bold
  modify_header(label ~ "**Variable**") %>%  # Customize the header
  modify_spanning_header(c(stat_1, stat_2) ~ "**Sex**") %>%
  as_gt()

# Save the gt table as a PNG file
gt::gtsave(gt_table1, filename = "images/table1.png", 
           path = getwd())
```


# DEMOGRAPHIC

### Name

```{r name}

```

### Sex

```{r sex}
exonerees %>% 
  group_by(Sex) %>%
  summarize(Count = n(), Proportion = Count / nrow(.),
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>%
  knitr::kable(col.names = c("Sex", "Count", "Proportion", 
                             "Years incarcerated"), digits = 2)
```

### Race

```{r race}
exonerees %>% 
  group_by(Race, Sex) %>%
  summarize(Count = n(), Proportion = Count / nrow(.),
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>%
  knitr::kable(col.names = c("Race", "Sex", "Count", "Proportion", 
                             "Years incarcerated"), digits = 2)
```

```{r}
exonerees %>% 
  group_by(Crime.Category) %>%
  summarize(Count = n(), Proportion = Count / nrow(.),
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>%
  knitr::kable(col.names = c("Crime Category", "Count", "Proportion", 
                             "Years incarcerated"), digits = 2)
```
```{r}
exonerees %>%
  group_by(Age.Group.at.Conviction) %>%
  summarize(Count = n(), Proportion = Count / nrow(.),
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>%
  knitr::kable(col.names = c("Age at conviction", "Count", "Proportion", 
                             "Years incarcerated"), digits = 2)
```



### Ages at events


```{r ages-at-events, message=FALSE}
# age groups at crime
# exonerees %>%
#   group_by(Age.Group.at.Crime) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   arrange(desc(Count)) %>%
#   knitr::kable(col.names = c("Age at Crime", "Count", "Proportion"), digits = 2)
# 
# exonerees %>% 
#   group_by(Age.Group.at.Crime, Age.Group.at.Release) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   arrange(desc(Count)) %>% filter(Count >= 95) %>%
#   knitr::kable(col.names = c("Age at crime", "Age at release", 
#                              "Count", "Proportion"), digits = 2)

# exonerees %>% 
#   group_by(Crime.Category, Age.Group.at.Crime, Age.Group.at.Release) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   arrange(desc(Count)) %>% filter(Count >= 95) %>%
#   knitr::kable(col.names = c("Crime category", "Age at crime", "Age at release", 
#                              "Count", "Proportion"), digits = 2)

### 

exonerees %>% 
  group_by(Age.Group.at.Conviction, Age.Group.at.Release) %>%
  summarize(Count = n(), Proportion = Count / nrow(.)) %>%
  arrange(desc(Count)) %>% filter(Count >= 95) %>%
  knitr::kable(col.names = c("Age at conviction", "Age at release", 
                             "Count", "Proportion"), digits = 2)

exonerees %>% 
  group_by(Crime.Category, Age.Group.at.Conviction, Age.Group.at.Release) %>%
  summarize(Count = n(), Proportion = Count / nrow(.)) %>%
  arrange(desc(Count)) %>% filter(Count >= 95) %>%
  knitr::kable(col.names = c("Crime category", "Age at conviction", "Age at release", 
                             "Count", "Proportion"), digits = 2)

### 

# exonerees %>% 
#   group_by(Race, Age.Group.at.Crime, Age.Group.at.Release) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.), 
#             Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
#   arrange(desc(Count)) %>% filter(Count >= 95) %>%
#   knitr::kable(col.names = c("Race", "Age at crime", "Age at release", 
#                              "Count", "Proportion", 
#                              "Total years incarcerated"), digits = 2)

# exonerees %>% 
#   group_by(Race, Age.Group.at.Crime, Age.Group.at.Release) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.), 
#             Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
#   arrange(desc(Total.Yrs.Incarcerated)) %>% filter(Count >= 95) %>%
#   knitr::kable(col.names = c("Race", "Age at crime", "Age at release", 
#                              "Count", "Proportion", 
#                              "Total years incarcerated"), digits = 2)

###

exonerees %>% 
  group_by(Race, Age.Group.at.Conviction, Age.Group.at.Release) %>%
  summarize(Count = n(), Proportion = Count / nrow(.), 
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>% 
  filter(Count >= 95) %>%
  knitr::kable(col.names = c("Race", "Age at conviction", "Age at release", 
                             "Count", "Proportion", 
                             "Years incarcerated"), digits = 2)

exonerees %>% 
  group_by(Race, Age.Group.at.Conviction, Age.Group.at.Release) %>%
  summarize(Count = n(), Proportion = Count / nrow(.), 
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Total.Yrs.Incarcerated)) %>% 
  filter(Count >= 95) %>%
  knitr::kable(col.names = c("Race", "Age at conviction", "Age at release", 
                             "Count", "Proportion", 
                             "Years incarcerated"), digits = 2)
```


```{r, message=FALSE}
# age at crime (by race)
# 
# exonerees %>% 
#   group_by(Age.Group) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   # arrange(desc(Count)) %>% filter(Count >= 100) %>%
#   knitr::kable(col.names = c("Age at Crime", "Count", "Proportion"), digits = 2)
# 
# 
# exonerees %>% 
#   mutate(Age.Group = cut(Age.at.Crime, breaks = age_groups, 
#                          labels = age_labels, include.lowest = TRUE)) %>%
#   group_by(Race, Age.Group) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   arrange(desc(Count)) %>% filter(Count >= 95) %>%
#   knitr::kable(col.names = c("Race", "Age at Crime", "Count", "Proportion"), 
#                digits = 2)

# # age at release (by race)
# 
# exonerees %>% 
#   mutate(Age.Group = cut(Age.at.Release, breaks = age_groups, 
#                          labels = age_labels, include.lowest = TRUE)) %>%
#   group_by(Age.Group) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   # arrange(desc(Count)) %>% filter(Count >= 100) %>%
#   knitr::kable(col.names = c("Age at Release", "Count", "Proportion"), digits = 2)
# 
# 
# exonerees %>% 
#   mutate(Age.Group = cut(Age.at.Release, breaks = age_groups, 
#                          labels = age_labels, include.lowest = TRUE)) %>%
#   group_by(Race, Age.Group) %>%
#   summarize(Count = n(), Proportion = Count / nrow(.)) %>%
#   arrange(desc(Count)) %>% filter(Count >= 95) %>%
#   knitr::kable(col.names = c("Race", "Age at Release", "Count", "Proportion"), 
#                digits = 2)


exonerees %>% 
  group_by(Race, Sex, Crime.Category) %>%
  summarize(Count = n(), Proportion = Count / nrow(.),
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>% filter(Count >= 95) %>%
  knitr::kable(col.names = c("Race", "Sex", "Crime category", 
                             "Count", "Proportion", "Years incarcerated"), 
               digits = 2)


exonerees %>% 
  group_by(Race, Sex, Crime.Category, Age.Group.at.Conviction) %>%
  summarize(Count = n(), Proportion = Count / nrow(.),
            Total.Yrs.Incarcerated = sum(Years.Incarcerated)) %>%
  arrange(desc(Count)) %>% filter(Count >= 95) %>%
  knitr::kable(col.names = c("Race", "Sex", "Crime category", 
                             "Age at conviction", "Count", "Proportion", 
                             "Years incarcerated"), 
               digits = 2)

```

## Derived demographic variables

### Age at conviction

```{r age-at-conviction}
exonerees %>% 
  select(Age.at.Conviction) %>%
  summary(digits = 2)
```

### Age at exoneration

```{r age-at-exoneration}
exonerees %>% 
  select(Age.at.Exoneration) %>%
  summary(digits = 2)
```

### Age at release

```{r age-at-release}
exonerees %>% 
  select(Age.at.Release) %>%
  summary(digits = 2)
```

# CONVICTION

### State

```{r state}

```

### County

```{r county}

```

### Year of crime

```{r year-of-crime}

```

### (Worst) crime

```{r worst-crime}

```

### Sentence

```{r sentence}

```

### Date of 1st conviction

```{r date-conviction}

```

### Conviction tags

## Derived conviction variables

### Year of 1st conviction

```{r year-conviction}

```

# EXONERATION

### Date of exoneration

```{r date-exoneration}

```

### Date of release

```{r date-release}

```

### Date of posting

```{r date-posting}

```

### Exoneration tags

## Derived exoneration variables

### Years served

```{r years-served}

```

### Year of exoneration

```{r year-exoneration}

```

### Year of release

```{r year-release}

```

### Year of posting

```{r year-posting}

```

**End of document**
